`methyvim`: Targeted and Model-free Analysis of Differential Methylation

### 2017-10-18 paper/paper_BiocF1000.Rmd
update+rebuild f1000 paper

---
title: '`methyvim`: an R package for Variable Importance Measures for
Differential Methylation Analysis'
author:
  - name: Nima S. Hejazi
    affiliation: 1
    orcid: 0000-0002-7127-2789
  - name: Rachael V. Phillips
    affiliation: 1
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: 1
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation: 1, 2
    orcid: TBA
affiliations:
  - name: Division of Biostatistics, University of California, Berkeley
    index: 1
  - name: Department of Statistics, University of California, Berkeley
    index: 2
abstract: |
  We present `methyvim`, a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with
  straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, variable importance, targeted minimum loss-based estimation
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging
[@libbrecht2015machine]. In this paper, we present an alternative such
statistical analysis approaches in the form of nonparametric estimation
procedures that provide simple and well interpretable statistical inference.
Inspired by concepts of counterfactual effects from the causal inference
literature and machine learning algorithms, we provide a computationally
efficient technique for obtaining targeted estimates of nonparametric _variable
importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
the False Discovery Rate (FDR) as if all sites were tested. Under
counterfactual assumptions, these estimators have asymptotic normal sampling
distributions and thus provide reliable inference. Counterfactuals are
"counter"-fact, defined in terms of an unrealistic yet ideal experiment. In the
context of DNA methylation and for a binary treatment, we define the
counterfactual outcomes as the Beta-value/M-value a CpG site would have if all
subjects were treated and the Beta-value/M-value a CpG site would have if all
subjects were not treated. Of course, these counterfactual outcomes are
impossible to observe but they are very helpful for us to asses the importance
of a CpG site relative to a subjects treatment/exposure and they can be
statistically estimated [@vdl2006statistical]. Our VIM definition uses the
concepts of counterfactuals and the framework for estimation is built on
_targeted maximum likelihood estimation_ or _targeted minimum loss-based
estimation_ (__TMLE__) [@vdl2006targeted]. This methodology targets the
individual importance of each CpG site by taking advantage of the prediction
power of machine learning algorithms. The mathematical properties of this
approach have been explored in [@tuglus2008targeted] and we focus on our
contribution to it's application, `methyvim`, an R package that make this
technique for differential methylation analysis easily accessible.

@@

We refer the reader to the package vignette, "`methyvim`: Variable Importance
Measures for Differential Methylation" hosted with the package on Bioconductor,
for a detailed description of the methods used in this package.
(NEED CITATION FOR VIGNETTE)

@@

```{r make-methytmle}
mtmle <- .methytmle(grs)
```
Additionally, a `GenomicRatioSet` can be created from a matrix with the
function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.

@@

```{r methyvim-pvals-both}
plot(methyvim_cancer_ate)
```
__Remark:__ The plots displayed above may also be generated separately by
explicitly setting the argument "type" to `plot.methytmle`. For a plot of the
raw p-values, specify `type = "raw_pvals"`, and for a plot of the FDR-corrected
p-values, specify `type = "fdr_pvals"`.

@@

Archived source code as at time of publication:

https://github.com/nhejazi/methyvim

Software license: MIT, copyright Nima S. Hejazi

>>>>

---
title: '`methyvim`: Targeted and Model-free Analysis of Differential Methylation
in R'
author:
  - name: Nima S. Hejazi
    affiliation: Division of Biostatistics, University of California, Berkeley
    orcid: 0000-0002-7127-2789
  - name: Rachael V. Phillips
    affiliation: Division of Biostatistics, University of California, Berkeley
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: Division of Biostatistics, University of California, Berkeley
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation: Division of Biostatistics, University of California, Berkeley
    orcid: TBA
abstract: |
  We present `methyvim`, a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with
  straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, statistical variable importance, targeted minimum loss-based
  estimation, machine learning
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging
[@libbrecht2015machine].

In this paper, we present an alternative such statistical analysis approaches in
the form of nonparametric estimation procedures that provide simple and well
interpretable statistical inference. Inspired by concepts of counterfactual
effects from the causal inference literature and machine learning algorithms, we
provide a computationally efficient technique for obtaining targeted estimates
of nonparametric _variable importance measures_ (__VIM__) for a pre-screened set
of CpGs, controlling for the False Discovery Rate (FDR) as if all sites were
tested. Under counterfactual assumptions, these estimators have asymptotic
normal sampling distributions and thus provide reliable inference.
Counterfactuals are "counter"-fact, defined in terms of an unrealistic yet ideal
experiment. In the context of DNA methylation and for a binary treatment, we
define the counterfactual outcomes as the observed methylation value (whether
Beta- or M-values) a CpG site would have if all subjects were treated and the
methylation value a CpG site would have if all subjects were not treated. Of
course, these counterfactual outcomes are impossible to observe but they are
very helpful for us to asses the importance of a CpG site relative to a subjects
treatment/exposure and they can be statistically estimated
[@vdl2006statistical]. Our VIM definition uses the concepts of counterfactuals
and the framework for estimation is built on _targeted maximum likelihood
estimation_ or _targeted minimum loss-based estimation_ (__TMLE__)
[@vdl2006targeted]. This methodology targets the individual importance of each
CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in
[@tuglus2008targeted] and we focus on our contribution to it's application,
`methyvim`, an R package that make this technique for differential methylation
analysis easily accessible.

@@

We refer the reader to the package vignette, "`methyvim`: Targeted Learning for
Differential Methylation Analysis," included in any distribution of the software
package, for further details.

@@

```{r make-methytmle}
library(methyvim)
mtmle <- .methytmle(grs)
```
Additionally, a `GenomicRatioSet` can be created from a matrix with the
function `makeGenomicRatioSetFromMatrix` provided by the `minfi` package.

@@

```{r methyvim-pvals-raw}
plot(methyvim_cancer_ate, type = "raw_pvals")
```

```{r methyvim-pvals-fdr}
plot(methyvim_cancer_ate, type = "fdr_pvals")
```

__Remark:__ The plots displayed above may also be generated as side-by-side
histograms in a single plot object. This is the default for the `plot` method
and may easily be invoked by specifying no additional arguments to the `plot`
function, unlike in the above.

@@

Archived source code as at time of
publication: https://github.com/nhejazi/methyvim

Software license: The MIT License, copyright Nima S. Hejazi
